Probabilistic Model of Bilateral Lymphatic Spread in Head and Neck Cancer

### 2024-08-27 _tex/manuscript.tex
Built site for gh-pages

These datasets allow us to investigate correlations between the
involvement of individual LNLs, or between risk factors and involvement.
In figure~\ref{fig-data-strat}, we have plotted the prevalence of each
contralateral LNL's involvement, stratified by T-category, number of
ipsilaterally involved LNLs, and whether the tumor extended over the
mid-sagittal line. A similar but more complete stratification is also
tabulated in the appendix in table~\ref{tbl-data-strat}.

@@

\caption{\label{fig-data-strat}Contralateral involvement stratified by
T-category (left panel), the number of metastatic LNLs ipsilaterally
(center panel), and whether the primary tumor extended over the
mid-sagittal line or was clearly lateralized (right panel).}

@@

The left panel in figure~\ref{fig-data-strat} indicates that T-category
is correlated with contralateral involvement (as it is with overal
involvement). This is simply because T-category may on average be
considered a surrogate for the time between onset of disease and
diagnosis. I.e., a patient with a T4 tumor was -- on average --
diagnosed later than a patient with a T1 tumor. Thus, the former did
have more time to develop metastases.

@@

Lastly, the right panel in figure~\ref{fig-data-strat} shows that
patients with a tumor crossing the mid-sagittal line show contralateral
involvement vastly more often compared to patients with clearly
lateralized tumors. This makes intuitive sense, because the lymphatic
system in the head and neck region is typically symmetric and thus no
major vessels cross the midline. Therefore, interstitial fluids from the
primary tumor -- which we assume to carry living malignant cells -- may
only reach the blind-ended lymphatic vessels in the contralateral neck
via short-ranged diffusion. Which in turn is only possible when the
primary tumor is close enough to the mid-sagittal line or crosses it.

The three risk factors for contralateral involvement, midline extension,
T-category, and ipsilateral involvement are correlated. For example,
midline extension was present in 48\% of advanced T-category tumors but
only in 7\% of early T-category tumors. The higher prevalence of
contralateral metastases in advanced T-category tumors may thus be
partially explained by a higher rate of midline extension in these
tumors. However, figure~\ref{fig-data-strat-uncorr} indicates that
advanced T-category and severe ipsilateral involvement do represent
additional risk factors. The figure considers only patients with
lateralized tumors not extending over the midline. For early T-category
and healthy ipsilateral levels I-V, only 1\% (1 out of 84 patients)
shows involvement of contralateral level II. This increases to 5\% (9
out of 165 patients) if ipsilateral level II is involved, to 15\% (9 out
of 60 patients) if ipsilateral levels II and III are involved, and to
23\% (6 out of 26 patients) for advanced T-category with ipsilateral
levels II and III involved.

\includegraphics{manuscript_files/mediabag/figures/fig_data_strat_uncorr.pdf}

\caption{\label{fig-data-strat-uncorr}Contralateral involvement
prevalence by LNL for lateralized tumors. Shown are four different
scenarios: 1) early T-category with an ipsilateral N0 neck (green), 2)
early T-category with ipsilaterally LNL II involved (blue), 3) late
T-category with ipsilateral LNL II metastatic (orange), and 4) early
T-category with ipsilateral both levels II and III involved (red). This
illustrates that even when controlling for the tumor's lateralization,
T-category and ipsilaeral involvement are predictive of contralateral
metastases.}

@@

It is possible to provide our model with multiple different diagnostic
modalities, each being characterized by different pairs of sensitivity
and specificity. However, we instead chose to combine them into a single
``consensus'' diagnosis before parameter inference, as introduced in
\textbf{?@sec-sec-data-consensus}. We opted for this because the
literature values of sensitivity and specificity
\citep{de_bondt_detection_2007, kyzas_18f-fluorodeoxyglucose_2008} of
imaging modalities like MRI and CT do not plausibly match some of our
observations: In the USZ cohort, 78\% of OPSCC patients where diagnosed
with ipsilateral LNL II involvement via diagnostic imaging. This is
virtually impossible with sensitivities around 80\% and specificities
lower than 100\%.

@@

\subsubsection{Prevalence of Midline
Extension}\label{prevalence-of-midline-extension}

Lastly, in figure~\ref{fig-model-prevalences-midext}, we plot the
prevalence of midline extension in the data versus our model's
prediction. It is obvious the model cannot match the large spread
between early and advanced T-category seen in the data. This is because
to achieve that, it would need to increase the advanced T-category
patient's prior distribution over diagnosis times and at the same time
reduce the probability of the tumor to cross the midline during a
time-step. But since the time-priors parameter is also coupled with the
spread probabilities among the LNLs, the model does not have that
freedom.

@@

>>>>

These datasets allow us to investigate correlations between the
involvement of individual LNLs, or between risk factors and involvement.
In figure~\ref{fig-data-strat}, we have plotted the prevalence of each
contralateral LNL's involvement, stratified by risk factors such as
T-category, number of ipsilaterally involved LNLs, and whether the tumor
extended over the mid-sagittal line. We also show in the fourth subfigre
that T-category and ipsilateral involvement are not just rpedictive of
contralateral involvement because they correlate with midline extension.
A similar but more complete stratification is also tabulated in the
appendix in table~\ref{tbl-data-strat}.

@@

\caption{\label{fig-data-strat}Contralateral involvement stratified by
T-category (top left panel), the number of metastatic LNLs ipsilaterally
(top right panel), and whether the primary tumor extended over the
mid-sagittal line or was clearly lateralized (bottom left panel).
Additionally, in the bottom right, we show that even for lateralized
tumors, T-category and ipsilateral involvement are correlated with
contralateral metastasis by comparing the contralateral involvement
prevalence for selected scenarios that vary in their T-category and
ipsilateral involvement extent.}

@@

The top left panel in figure~\ref{fig-data-strat} indicates that
T-category is correlated with contralateral involvement (as it is with
overal involvement). This is simply because T-category may on average be
considered a surrogate for the time between onset of disease and
diagnosis. I.e., a patient with a T4 tumor was -- on average --
diagnosed later than a patient with a T1 tumor. Thus, the former did
have more time to develop metastases.

@@

The bottom left panel in figure~\ref{fig-data-strat} shows that patients
with a tumor crossing the mid-sagittal line show contralateral
involvement vastly more often compared to patients with clearly
lateralized tumors. This makes intuitive sense, because the lymphatic
system in the head and neck region is typically symmetric and thus no
major vessels cross the midline. Therefore, interstitial fluids from the
primary tumor -- which we assume to carry living malignant cells -- may
only reach the blind-ended lymphatic vessels in the contralateral neck
via short-ranged diffusion. Which in turn is only possible when the
primary tumor is close enough to the mid-sagittal line or crosses it.

The three risk factors for contralateral involvement, midline extension,
T-category, and ipsilateral involvement are correlated. For example,
midline extension was present in 48\% of advanced T-category tumors but
only in 7\% of early T-category tumors. The higher prevalence of
contralateral metastases in advanced T-category tumors may thus be
partially explained by a higher rate of midline extension in these
tumors. However, the bottom right panel of figure~\ref{fig-data-strat}
indicates that advanced T-category and severe ipsilateral involvement do
represent additional risk factors. The figure considers only patients
with lateralized tumors not extending over the midline. For early
T-category and healthy ipsilateral levels I-V, only 1\% (1 out of 84
patients) shows involvement of contralateral level II. This increases to
5\% (9 out of 165 patients) if ipsilateral level II is involved, to 15\%
(9 out of 60 patients) if ipsilateral levels II and III are involved,
and to 23\% (6 out of 26 patients) for advanced T-category with
ipsilateral levels II and III involved.

@@

It is possible to provide our model with multiple different diagnostic
modalities, each being characterized by different pairs of sensitivity
and specificity. However, we instead chose to combine them into a single
``consensus'' diagnosis before parameter inference, as introduced in
section~\ref{sec-data-consensus}. We opted for this because the
literature values of sensitivity and specificity
\citep{de_bondt_detection_2007, kyzas_18f-fluorodeoxyglucose_2008} of
imaging modalities like MRI and CT do not plausibly match some of our
observations: In the USZ cohort, 78\% of OPSCC patients where diagnosed
with ipsilateral LNL II involvement via diagnostic imaging. This is
virtually impossible with sensitivities around 80\% and specificities
lower than 100\%.

@@

@@

\section{Prevalence of Midline
Extension}\label{prevalence-of-midline-extension}

In figure~\ref{fig-model-prevalences-midext}, we plot the prevalence of
midline extension in the data versus our model's prediction. It is
obvious the model cannot match the large spread between early and
advanced T-category seen in the data. This is because to achieve that,
it would need to increase the advanced T-category patient's prior
distribution over diagnosis times and at the same time reduce the
probability of the tumor to cross the midline during a time-step. But
since the time-priors parameter is also coupled with the spread
probabilities among the LNLs, the model does not have that freedom.
